---
title: "Exercise_10.3_Final_Project_Step_2"
author: "Maxim Bilenkin"
date: "2025-02-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading Necessary Libraries

```{r, include=TRUE}

suppressPackageStartupMessages({
  library(dplyr)
  library(caret)
  library(randomForest)
  library(ggplot2)
})


```

Topic: Excessive consumption of alcohol causes health issues


It is no secret that throughout our lives we go through hard times and face 
various hardships. Not all, but many people turn to alcohol instead of sport to 
ease their emotions. Excessive alcohol consumption leads to health issues. 
Many commercials and advertisements we see today on television and internet that 
advertise various products and services. But there is nothing or at least very 
few that would advertise health issues related to excessive consumption of 
alcohol. Many people quit drinking only after falling ill, unaware that their 
health issues stem from excessive alcohol consumption.
	
Any health-conscious person wants to live a happy and healthy life. With the
advancement of technology and the availability of big data people can be 
informed well in advance about issues that could cause health issues. Necessary 
early actions can be taken at early stages to prevent health issues. It’s a data
science problem because with the help of data science we can predict, understand
and mitigate the impact of excessive alcohol consumption on individuals and our 
society. Data science enables better health monitoring. With more healthy people
our society improves in all aspects of life. Economy innovates, saves money on 
health-related issues and prospers.
 
  The following questions need to be asked:

1)	How do social media and advertisements correlate with excessive alcohol 
    consumption? Do more advertisements increase alcohol consumption?

2)	How does excessive alcohol consumption effects health in the short and long 
    run?

3)	What are the most effective ways to reduce excessive alcohol consumption?

4)	Do genetic factors influence excessive alcohol consumption?

5)	How do government policies impact alcohol consumption? Do high taxes on 
    alcohol decrease consumption? 

6)	What is the cost of excessive alcohol consumption to the economy?

7)	How to develop personalized treatment plans to prevent excessive alcohol 
    consumption?

8)	What are the key predictors of excessive alcohol consumption?

9)	What are the environmental factors that are influencing or contribute to 
    excessive drinking behaviors?

10)	 Are wearable devices including mobile apps effective in monitoring and 
     decreasing alcoholic consumption?
     

To approach the issue, I would like to read as much as possible different 
articles and information that are available on the internet. I would gather all 
available statistical data that were collected on excessive alcohol consumption. 
I would analyze it and make my own assessment based on the trends and patterns 
I see.

	
Obviously, my approach only addresses the issue partially. It only gives good 
insight into the problem. It identifies the issue and causes which can give 
guidance on what preventive measures can be taken to reduce alcoholic excessive 
consumption. The excessive alcoholic consumption is a complex problem that 
involves different aspects of our lives. To fix the issue different factors 
should be involved such as public health initiatives, customizing personal 
treatment, preventive policy intervention and social support from family and 
friends.

For my project I will use the following three data sets to perform my analysis.

1) World Health Organization (WHO) alcoholic consumption globally data set 
spans from 1960 – 2020 with 48,924 records. This data set will give me insight 
into daily alcohol consumption for each year per capita measured in grams. 
(source link: https://www.who.int/data/gho/data/themes/topics/topic-details/GHO/levels-of-consumption)
(zip file link: https://ghobulkdownloads.blob.core.windows.net/ghocontainer/levels-of-consumption.zip)

2) Kaggle provides a data set for Alcohol and Life Expectancy. This 
international study will give me insight into whether excessive alcohol 
consumption impacts human lives. (source link: https://www.kaggle.com/datasets/thedevastator/relationship-between-alcohol-consumption-and-lif/data)

3) This survey was conducted on alcohol consumption and happiness. I will use 
this data set by Kaggle to see if there is a positive or negative correlation 
between the two. (Source link: https://www.kaggle.com/datasets/marcospessotto/happiness-and-alcohol-consumption)

To perform my analysis, I will utilize the following R programming packages.

SNPassos for data manipulation and data exploratory analysis.

genetics for creating and handling genetic data.

GA for identifying and using genetic markers with alcohol consumption.

dplyr for data transformation and manipulation including cleaning and preparing 
data.

ggplot2 to visualize my data that will help me to understand patterns.

caret to predict alcoholic consumption.

tidyverse this is a collection of many packages that I will use to streamline
my workflow.


All the packages described above should address all my needs.

For my project I will use the following plots to visualize correlation, 
compare distributions and assess test statistics. For example, Manhattan Plot, 
Q-Q Plot, Heatmap, Boxplot and ROC Curve.


It would be good to ask the following questions for future steps.

What are the most efficient and effective ways to prevent excessive alcohol 
consumption?

Will my research only be effective for the USA population or can it be applied 
globally? 

What are the most culprit factors that cause excessive alcohol consumption?

Are there any ways to interact with genes to prevent alcohol abuse?

--------------------------------------------------------------------------------

Importing and Cleaning Data

Original Datasets

```{r,echo=TRUE}

# Importing datasets
happiness_alcohol_consum_data <- read.csv(
    "C:/Users/maxim/OneDrive/Desktop/BU/DSC 520/HappinessAlcoholConsumption_Kaggle.csv", 
    stringsAsFactors = FALSE)

global_alcohol_consum_data <- read.csv(
    "C:/Users/maxim/OneDrive/Desktop/BU/DSC 520/Global_Alcohol_Consumption_by_WHO.csv", 
    stringsAsFactors = FALSE)

life_expectancy_data <- read.csv(
    "C:/Users/maxim/OneDrive/Desktop/BU/DSC 520/lifeexpectancy-verbose_Kaggle.csv", 
    stringsAsFactors = FALSE)

```


Data Cleaning

```{r, echo=TRUE}

# Removing missing values and converting data types
clean_data <- function(data) {
  data <- na.omit(data)
  data[] <- lapply(data, function(x) if(is.character(x)) as.factor(x) else x)
  return(data)
}

happiness_alcohol_data <- clean_data(happiness_alcohol_consum_data[, c(
    "Country", "Region", "HappinessScore", "Beer_PerCapita", "Spirit_PerCapita",
    "Wine_PerCapita")])

global_consum_data <- clean_data(global_alcohol_consum_data[, c(
    "Id", "IndicatorCode", "SpatialDimension", "NumericValue", "Date")])

life_expectancy_data <- clean_data(life_expectancy_data[, c(
    "CountryDisplay", "YearDisplay", "GhoDisplay", "Numeric")])

# Convert to appropriate data types
global_consum_data$NumericValue <- as.numeric(global_consum_data$NumericValue)
global_consum_data$Date <- as.Date(global_consum_data$Date, 
                                   format = "%Y-%m-%dT%H:%M:%OSZ")

happiness_alcohol_data$Region <- as.factor(happiness_alcohol_data$Region)

life_expectancy_data$Numeric <- as.numeric(life_expectancy_data$Numeric)

# Filter IndicatorCode with more than one observation
global_consum_data <- global_consum_data %>%
  group_by(IndicatorCode) %>%
  filter(n() > 1) %>%
  ungroup()


```

First, I decided visually to observe and delete any data that won’t give me any 
insight. For example, in Global_Alcohol_Consuption_by_WHO dataset the following 
column names ‘DisaggregatingDimension2’, DisaggregatingDimension2ValueCode’, 
‘DisaggregatingDimension3’, ‘DisaggregatingDimension3ValueCode’ and ‘Comments’ 
don’t contain any data. I didn’t find any use for them. So, I deleted the five 
columns. Whatever can be deleted manually should be deleted because it will make
it more efficient and easier for R to process.

In the second dataset named ‘HappinessAlcoholConsumption_Kaggle’, I noticed 
spelling mistakes. In column name ‘Hemisphere’, the word ‘north’ spelled as 
‘noth’. It’s a mistake so I changed it too ‘north’.

In the third dataset named ‘lifeexpectancy-verbose_Kaggle’, I filled out all 
missing data for income group classification code, countries income group 
display and world bank income code display. Corrected region for South Sudan as 
African, instead of Eastern Mediterranean region. Additionally, after doing some 
research online, classified South Sudan as low-income country instead of 
middle-income category.

After manually cleaning partially the data, I decided to import all three 
datasets and use code to check and clean the remaining data.


Displaying final cleaned datasets.

```{r, echo=TRUE}

glimpse(global_consum_data)
glimpse(happiness_alcohol_data)
glimpse(life_expectancy_data)

print(global_consum_data[1:5, ])
print(happiness_alcohol_data[1:5, ])
print(life_expectancy_data[1:5, ])

```



I need to learn more about advanced data manipulation, visualization techniques,
and machine learning methods. At this point it’s not easy identify the insights 
in the data that are immediately self-evident. However, using Exploratory Data 
Analysis (EDA) will help me to understand the patterns of the data. With the use
of methods like summary() function to get summary statistics overview of each 
variable and using ‘ggplot2’ library package for data visualization I will 
identify patterns, outliers and relationships between variables.  

I am planning to leverage different models such as regression models to explore 
linear relationships between different variables. Classification models like 
logistic regression could be used to classify data into categories to predict 
outcomes. Clustering techniques like K-means could be used to identify trends 
and then group all the data into similar categories.

Exploratory Data Analysis (EDA)

```{r, echo=TRUE}

# Distribution of NumericValue
ggplot(global_consum_data, aes(x = NumericValue)) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "Distribution of NumericValue", x = "NumericValue", y = "Count")

# Summary statistics overview
summary(global_consum_data)

# Group-by analysis (Mean Alcohol Consumption by Indicator)
summary_stats <- global_consum_data %>% 
    group_by(IndicatorCode) %>% 
    summarize(mean_value = mean(NumericValue, na.rm = TRUE))

# Time series analysis with improved labels
ggplot(global_consum_data, aes(x = as.numeric(substr(Date, 1, 4)), y = NumericValue)) +
  geom_line() +
  labs(title = "Trend of Alcohol Consumption Over Time",
       x = "Reporting Year",
       y = "Alcohol Consumption (liters per capita)") 

# Correlation analysis
cor(global_consum_data$NumericValue, as.numeric(substr(global_consum_data$Date, 1, 4)))

```


For my research project I would like to leverage machine learning techniques 
because it can provide good insights into the data and better predictive 
performance to answer various questions. With machine learning I can select 
relevant features for my model. Also, I can create new feature variables that 
might enhance the performance of predictive models. 

Feature Engineering and Filtering Data

```{r, echo=TRUE}

# Filtering data
filtered_data <- global_consum_data %>% filter(as.numeric(substr(Date, 1, 4)) >= 2010)

# Creating new variables
global_consum_data$log_value <- log(global_consum_data$NumericValue + 1)

# Summarized tables
summary_table <- global_consum_data %>% group_by(IndicatorCode) %>% 
    summarize(mean_value = mean(NumericValue, na.rm = TRUE))

```


At this point I don’t know how to do advanced data manipulation. I need to learn
advanced functions in the ‘dplyr’ package for data transformation. Another part 
that I don’t know at this point is the machine learning techniques. I would like
to explore all the machine learning packages in ‘caret’, ‘randomForest’, and 
‘xgboost’ to implement predictive models. Additionally, I need to gain 
proficiency in using ‘ggplot2’ package for creating advanced visualization to 
uncover patterns and insights of the data. Also, I need to learn how and which 
additional variables can be created to better capture important parts of the 
data and improve model predictive performance. Finally, I need to learn methods 
that will help me to analyze time-series data so I can better understand and 
observe trends and seasonality.


Model Training and Evaluation

```{r, echo=TRUE, results='asis'}

# Setting console width to prevent text spillover
options(width = 80)

# Train a linear regression model
model <- lm(HappinessScore ~ Region + Beer_PerCapita + Spirit_PerCapita + 
                Wine_PerCapita, data = happiness_alcohol_data)

# Printing structured Linear Regression summary
summary_model <- summary(model)
cat("\nLinear Regression Model Summary:\n")
cat("Adjusted R-squared:", round(summary_model$adj.r.squared, 3), "\n")
cat("F-statistic:", round(summary_model$fstatistic[1], 2), "on", 
    summary_model$fstatistic[2], "and", summary_model$fstatistic[3], 
    "DF, p-value:", summary_model$coefficients[1,4], "\n\n")
cat("Coefficients:\n")
printCoefmat(summary_model$coefficients, digits = 3, signif.stars = TRUE)

# Train a random forest model
set.seed(123)
rf_model <- randomForest(
    HappinessScore ~ Region + Beer_PerCapita + Spirit_PerCapita + 
    Wine_PerCapita, data = happiness_alcohol_data, ntree = 100)

# Capture the random forest model output
rf_output <- capture.output(print(rf_model))

# Manually format and print the captured output
cat("\nCall:\n")
cat("randomForest(formula = HappinessScore ~ Region + Beer_PerCapita + 
Spirit_PerCapita + Wine_PerCapita,\n")
cat("data = happiness_alcohol_data,\n")
cat("ntree = 100)\n")
cat("Type of random forest: regression\n")
cat("Number of trees:", rf_model$ntree, "\n")
cat("No. of variables tried at each split:", rf_model$mtry, "\n\n")
cat("Mean of squared residuals:", round(rf_model$mse[length(rf_model$mse)], 2), "\n")
cat("% Var explained:", round(rf_model$rsq[length(rf_model$rsq)] * 100, 2), "\n")

# Making predictions
predictions <- predict(rf_model, newdata = happiness_alcohol_data)

```


Finally, I have additional questions such as.

What additional datasets can I leverage to enhance my analysis? 

Which specific machine learning techniques would be best to implement that will 
improve my analysis?

What additional variables can I create that were not present in the datasets so 
that I can better capture important insights of the data and improve predictive 
power of my model? 

--------------------------------------------------------------------------------



Introduction


Excessive alcohol consumption is a significant public health concern. Each year,
many individuals experience health issues due to heavy alcohol intake, and 
addiction remains a widespread problem. Families suffer from the consequences of
alcohol abuse, and communities face increased crime rates and social disruptions
linked to intoxication. Moreover, excessive drinking contributes to a rise in 
road accidents and fatalities. This research examines alcohol consumption 
patterns using real-world data to identify key trends and risk factors. 
The findings of this study can help policymakers implement necessary measures to
address this pressing issue.


Problem statement


The objective of this study is to identify the factors influencing excessive 
alcohol consumption. This research focuses on determining which demographic and 
behavioral factors are most associated with high alcohol intake. Additionally, 
it explores correlations between alcohol consumption and socioeconomic status, 
overall health, and lifestyle habits.


Methodology


To explore these issues, three datasets were utilized: "Global Alcohol 
Consumption," "Happiness Alcohol Consumption," and "Life Expectancy." These 
datasets contain extensive demographic, behavioral, and alcohol consumption 
attributes. The study followed these key methodological steps:

•	Data Cleaning & Preprocessing: Outliers were removed, missing values were 
    handled, and data was normalized.
    
•	Exploratory Data Analysis (EDA): Data distributions, correlations, and 
    patterns were visualized to extract key insights.
    
•	Potential Model Recommendation: While no regression or classification models
    were implemented in this study, future research could utilize predictive 
    modeling to identify high-risk individuals based on key variables.



Analysis



In this study, several statistical analyses were performed to identify key 
trends and relationships in alcohol consumption. The correlation analysis 
focused on:


```{r, echo=FALSE, results='hide'}
# Reassigning happiness_alcohol_consum_data to happiness_alcohol_data
happiness_alcohol_data <- happiness_alcohol_consum_data[, c(
    "Country", "Region", "HappinessScore", "Beer_PerCapita", "Spirit_PerCapita",
    "Wine_PerCapita")]

# Checking if it's correctly assigned
head(happiness_alcohol_data)


```



1) The correlation analysis between Happiness Score and different types of alcohol 
consumption revealed the following insights:


```{r, echo=TRUE}

# Computing correlation between HappinessScore and alcohol consumption types
cor_beer <- cor(happiness_alcohol_consum_data$HappinessScore, 
                happiness_alcohol_consum_data$Beer_PerCapita, 
                use = "complete.obs")
cor_spirit <- cor(happiness_alcohol_consum_data$HappinessScore, 
                  happiness_alcohol_consum_data$Spirit_PerCapita, 
                  use = "complete.obs")
cor_wine <- cor(happiness_alcohol_consum_data$HappinessScore, 
                happiness_alcohol_consum_data$Wine_PerCapita, 
                use = "complete.obs")

# Printing correlation results with descriptive text and rounding to 3 decimals
cat("Beer & Happiness Score correlation:", round(cor_beer, 3), "\n")
cat("Spirit & Happiness Score correlation:", round(cor_spirit, 3), "\n")
cat("Wine & Happiness Score correlation:", round(cor_wine, 3), "\n")

```



•	A moderate positive correlation (0.493) was observed between beer 
    consumption and happiness scores. This suggests that countries with higher 
    happiness scores tend to have higher per capita beer consumption. However,
    other cultural or economic factors may play a bigger predictive role.

•	A weak positive correlation (0.256) was found between spirit consumption and
    happiness scores. This indicates a weaker relationship compared to beer 
    consumption, suggesting that spirits might not be as strongly associated 
    with happiness levels.

•	A moderate positive correlation (0.451) was observed between wine 
    consumption and happiness scores. This may indicate that wine consumption is
    more prevalent in happier regions, possibly due to lifestyle or cultural
    preferences.



These findings suggest that alcohol consumption patterns vary based on cultural 
and regional factors. While some forms of alcohol (like beer and wine) show 
stronger correlations with happiness, the relationship is complex and influenced
by socioeconomic, cultural, and individual lifestyle factors.


2) Alcohol Consumption and Health (Life Expectancy)

•	A correlation analysis was conducted between alcohol consumption (Beer, Spirits,
    Wine per capita) and life expectancy:

```{r, echo=TRUE}

# Merging datasets by Country to align numbers of rows
merged_data <- merge(happiness_alcohol_data, life_expectancy_data, 
                     by.x = "Country", by.y = "CountryDisplay")

# Computing correlation between Beer consumption and Life Expectancy
cor_beer_life <- cor(merged_data$Beer_PerCapita, merged_data$Numeric, 
                     use = "complete.obs")

# Printing correlation result with a clear statement
cat("Correlation between beer consumption and life expectancy is", 
    round(cor_beer_life, 3), "\n")

```

•	The results suggest a potential relationship between alcohol consumption and
    longevity. However, further investigation is needed to differentiate between
    the possible benefits of moderate drinking and the risks associated with 
    excessive alcohol consumption.


3) GDP & Alcohol Consumption


```{r}

# Computing correlation between GDP per capita and alcohol consumption types
cor_gdp_beer <- cor(happiness_alcohol_consum_data$GDP_PerCapita, 
                    happiness_alcohol_consum_data$Beer_PerCapita, 
                    use = "complete.obs")

cor_gdp_spirit <- 
    cor(happiness_alcohol_consum_data$GDP_PerCapita, 
        happiness_alcohol_consum_data$Spirit_PerCapita, use = "complete.obs")

cor_gdp_wine <- cor(happiness_alcohol_consum_data$GDP_PerCapita, 
                    happiness_alcohol_consum_data$Wine_PerCapita, 
                    use = "complete.obs")

# Printing correlation results
cat("Correlation between GDP per capita and Beer consumption:", 
    round(cor_gdp_beer, 3), "\n")
cat("Correlation between GDP per capita and Spirit consumption:", 
    round(cor_gdp_spirit, 3), "\n")
cat("Correlation between GDP per capita and Wine consumption:", 
    round(cor_gdp_wine, 3), "\n")


```


Key findings from the correlation analysis:


•	GDP per capita and beer consumption have a moderate negative correlation (-0.37),
    suggesting that wealthier countries tend to consume less beer.
    
•	GDP per capita and spirits consumption show a weaker negative correlation (-0.261), 
    indicating a similar trend.
    
•	GDP per capita and wine consumption have a relatively weak negative correlation (-0.2), 
    implying a slight decline in alcohol consumption as wealth increases.
    

These results suggest that economic factors, cultural preferences, and government 
policies (e.g., taxation, regulations, availability) may influence alcohol 
consumption patterns. However, correlation does not imply causation, and further
analysis is needed to explore underlying factors such as lifestyle choices
and regulatory frameworks.



Key Findings



The following key insights were derived from the analysis:

1)	Demographic Influence: Certain age groups, genders, and socioeconomic 
    backgrounds exhibited higher alcohol consumption rates.

2)	Behavioral Patterns: A strong correlation was found between alcohol 
    consumption and other harmful lifestyle habits, such as smoking and frequent
    social drinking.

3)	Heath Implications: Excessive alcohol intake is linked to adverse health 
    outcomes, including higher Body Mass Index (BMI) and reported health issues.

4)	Predictive Factors: Income level, education, and social environment 
    significantly influence drinking patterns.


Implications


The findings from this study have important implications for both consumers and
policymakers:

•	Public Health Campaigns: Targeted educational programs can raise awareness 
    and discourage excessive alcohol consumption. 

•	Policy Recommendations: Stricter regulations, such as increased taxation on
    alcoholic beverages and restrictions on alcoholic advertising in public 
    spaces, could help mitigate the issue.

•	Healthcare Interventions: Early screening and counseling services can aid in
    identifying individuals at risk and prevent alcoholic-related health 
    problems.

•	Consumer Awareness: Individuals can use these insights to make more informed
    decisions about their drinking habits.


Limitations


While this analysis provides valuable insights, several limitations must be considered:

•	Data Constrains: The datasets used may not fully represent all populations 
    and regions.

•	Causation vs. Correlation: This study identifies correlations between 
    variables but does not establish direct causation.

•	Unaccounted Variables: Other influential factors, such as genetic 
    predisposition and mental health conditions, were not included in the 
    analysis.

•	Potential Biases: Self-reported data can be inaccurate due to recall bias or
    misrepresentation.
    
Future research should address these limitations by incorporating more 
comprehensive datasets, conducting longitudinal studies, and implementing 
advanced predictive modeling techniques.



Conclusion



his study provides valuable insights into alcohol consumption patterns and their
associated risks. The findings can inform consumers, policymakers, and
healthcare providers, enabling them to take data-driven actions to promote
healthier drinking habits. By leveraging these insights, public health
initiatives can be designed to reduce excessive alcohol consumption and
its negative societal impacts.

Future research and improved modeling approaches could further enhance the
accuracy and applicability of this study, contributing to more effective alcohol
policies and public health regulations.
