---
title: "ASSIGNMENT (4.2 Exercise)"
author: "Maxim Bilenkin"
date: "2024-12-17"
output: pdf_document
---

# Assignment: ASSIGNMENT (4.2 Exercise)
# Name: Bilenkin, Maxim
# Date: 2024-12-17

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "markup")
options(repos = c(CRAN = "https://cran.r-project.org"))

```{r, message=FALSE, warning=FALSE}
# Loading the necessary packages
library(readr)
library(dplyr)

```{r, message=FALSE, warning=FALSE}
# Importing data file with scores
scores_data_file <- read.csv("C:\\Users\\maxim\\Downloads\\scores.csv")

# Displaying the file with all the data including fields to understand it
str(scores_data_file)
summary(scores_data_file)
colnames(scores_data_file)
head(scores_data_file)

```{r, message=FALSE, warning=FALSE}
# Separating the sections
sports_section <- scores_data_file %>% filter(Section == "Sports")
regular_section <- scores_data_file %>% filter(Section == "Regular")

```{r, message=FALSE, warning=FALSE}
summary(regular_section$Score)

```{r, message=FALSE, warning=FALSE}
# Summary for the Sports section.
summary(sports_section$Score)

```{r, message=FALSE, warning=FALSE}
sum(is.na(regular_section$Score))

```{r, message=FALSE, warning=FALSE}
head(regular_section)

```{r, message=FALSE, warning=FALSE}
# Removing rows with NA values in the Score column
cleaned_regular_section <- regular_section %>% filter(!is.na(Score))

# Descriptive statistics for the cleaned variety section
regular_variety_summary <- cleaned_regular_section %>% summarize(
  mean_score = mean(Score),
  standard_deviation_score = sd(Score)
)

# Printing cleaned variety summary
print(regular_variety_summary)

```{r, message=FALSE, warning=FALSE}
# Descriptive statistics for the sports section
sports_summary <- sports_section %>% summarize(
  mean_score = mean(Score),
  standard_deviation_score = sd(Score)
)

# Printing sports summary
print(sports_summary)

```{r, message=FALSE, warning=FALSE}
# Descriptive statistics for the variety section
variety_summary <- regular_section %>% summarize(
  mean_score = mean(Score),
  standard_deviation_score = sd(Score)
)

# Printing variety summary
print(variety_summary)

# 1) What are the observational units in this study?

# Answer: The observational units in this study are the student representing
#         a data point. Each student has it's own fields such as "Count", 
#         "Score" and "Section".

# 2) Identify the variables mentioned in the narrative paragraph and determine 
# which are categorical and quantitative?

# Answer: There are three variables, "Count", "Score" and "Section". Both, Count
#         and Score represent numbers and are therefore quantitative variables. 
#         On the other hand, Section represents a category of each student 
#         making it a categorical variable because it describes the section each
#         student belongs to.

# 3) Create one variable to hold a subset of your data set that contains only 
#    the Regular Section and one variable for the Sports Section.

# Answer:

# Creating a subset for Regular Section.
regular_section <- scores_data_file %>% filter(Section == "Regular")
head(regular_section)

# Creating a subset for Sports Section.    
sports_section <- scores_data_file %>% filter(Section == "Sports")
head(sports_section)

# 4) Use the Plot function to plot each Sections scores and the number of 
#    students achieving that score. Use additional Plot Arguments to label the 
#    graph and give each axis an appropriate label. Once you have produced your 
#    Plots answer the following questions:

# Answer:

# Creating a subset for Regular Section.
regular_section <- scores_data_file %>% filter(Section == "Regular")

# Creating a subset for Sports Section.
sports_section <- scores_data_file %>% filter(Section == "Sports")

# Defining a common x-axis ranges. 
x_range <- range(c(regular_section$Score, sports_section$Score))

# Plot for the Regular Section
plot(regular_section$Score, regular_section$Count, 
     main = "Scores vs. Number of Students (Regular Section)",
     xlab = "Scores", ylab = "Number of Students", 
     col = "blue", pch = 19, xlim = x_range)

# Plot for the Sports Section
plot(sports_section$Score, sports_section$Count, 
     main = "Scores vs. Number of Students (Sports Section)",
     xlab = "Scores", ylab = "Number of Students", 
     col = "red", pch = 19, xlim = x_range)

# a. Comparing and contrasting the point distributions between the two 
#    section, looking at both tendency and consistency: Can you say that one 
#    section tended to score more points than the other? Justify and explain
#    your answer.

# Answer: Based on the descriptive statistics we can see that Regular point 
#         distribution has a mean of 327.63 and Sports point distribution has a 
#         mean of 307.37. Since, Regular mean is higher than Sports mean. It 
#         means that the Regular students scoring more points because its 
#         tendency distribution is higher. Also, we can see that the standard 
#         deviation of Regular is 33.27 which is lower than standard deviation 
#         of 58.03 for Sports distribution. This means that the Sports 
#         distribution points spread more from the mean. At the same time it 
#         means that that the Regular distribution points spread less from the 
#         mean and it is more consistent.

# b. Did every student in one section score more points than every student in 
#    the other section? If not, explain what a statistical tendency means in 
#    this context.

# Answer: Not every student in one section scored more points than in another 
#         section. We can easily and quickly see by taking the minimum and 
#         maximum number for each distribution points. For Regular min and max 
#         are 265 and 380 respectively. And for Sports min and max are 200 and 
#         395 respectively. We can see clearly that some points on Sports 
#         distribution have scores for students that are higher with 395 vs. 380
#         for Regular students. However, statistical tendency means that on 
#         average students for Regular distribution score higher.

# c. What could be one additional variable that was not mentioned in the 
#    narrative that could be influencing the point distributions between the two
#    sections?

# Answer: Extra participation in some sport curriculum could be the additional 
#         variable that probably influencing students performance on scores. We
#         could say that students in regular section have more time to spend and
#         concentrate on their studies and homework than students from sport
#         section. This could be a good explanation why scoring is higher for 
#         regular students on average.